```{=html}
<style>
details {
  background-color: #F8F9FA;
  padding: 5px;
  border: 1px solid #ddd;
  border-radius: 5px;
  margin: 10px 0;
}
details summary {
  cursor: pointer;
}
</style>
```

# Regression II

Quantifying relationships between variables is an important skill 
in business analytics. The regression line, representing the best 
linear fit between two variables, enables businesses to make 
predictions, identify trends, and optimize strategies using 
data-driven insights. Understanding how to calculate and interpret 
a regression line provides the foundation for analyzing business 
relationships, such as the effect of advertising on sales or 
customer satisfaction on revenue. Below, we delve into generating 
and analyzing a regression line.

## The Regression Line

The regression line is calculated to minimize the average distance 
(or errors) between the line and the observed data points. It is 
defined by two key components: a **slope** ($\hat{\beta}$) and an 
**intercept** ($\hat{\alpha}$). Mathematically, the regression line 
is expressed as:

::: {.callout-formula}
**THE REGRESSION LINE**
$$\hat{y_i}=\hat{\alpha}+\hat{\beta}x_i$$
where $\hat{y_i}$ are the predicted values of $y$ given $x_i$, 
$\hat{\alpha}$ is the intercept, and $\hat{\beta}$ is the slope.
:::

The **slope** determines the steepness of the line and quantifies 
how much a unit increase in $x$ changes $y$ on average. It is 
calculated using the covariance between $y$ and $x$ and the 
variance of $x$:

::: {.callout-formula}
**THE SLOPE**
$$\hat{\beta}= \frac{s_{xy}}{s_{x}^2}$$
where $s_{xy}$ is the sample covariance between $x$ and $y$, and 
$s_x^2$ is the sample variance of $x$.
:::

The **intercept** determines where the line crosses the $y$ axis — 
the predicted value of $y$ when $x$ is zero. Once the slope is 
estimated, the intercept is calculated by:

::: {.callout-formula}
**THE INTERCEPT**
$$\hat{\alpha}=\bar{y}-\hat{\beta}\bar{x}$$
where $\bar{y}$ is the mean of $y$ and $\bar{x}$ is the mean of $x$.
:::

::: {.callout-example}
**Example:** Let's examine a data set on Price and Advertisement. 
In general, one expects that when a company advertises, it can 
convince consumers to pay more for their product. Below is the data:

| Advertisement (x) | Price (y) |
|:-----------------:|:---------:|
|         2         |     7     |
|         1         |     3     |
|         3         |     8     |
|         4         |    10     |

The data shows a clear direct relationship between advertisement 
and price. Regression allows us to answer two key questions:

1. **Effectiveness:** How much can we increase the price for every 
additional dollar spent on advertisement?
2. **Prediction:** What is the predicted price if we have a budget 
of $6$ for advertisement?

**Step 1 — Calculate the slope:** Given that the covariance is 
$s_{xy}=3.67$ and the variance of $x$ is $s_x^2=1.67$:

$$\hat{\beta}= \frac{3.67}{1.67}=2.2$$

For every additional dollar spent on advertisement, price increases 
by $2.2$ on average.

**Step 2 — Calculate the intercept:** Since $\bar{x}=2.5$ and 
$\bar{y}=7$:

$$\hat{\alpha}=7-2.2(2.5)=1.5$$

If we do not advertise, the predicted price is $1.5$.

**Step 3 — Write the regression line:**

$$\hat{y_i}=1.5+2.2x_i$$

**Step 4 — Make a prediction:** With an advertisement budget of $6$:

$$\hat{y_i}=1.5+2.2(6)=14.7$$

In conclusion, for every dollar spent on advertisement the price 
increases by $2.2$, and with a budget of $6$ the predicted price 
is $14.7$.
:::

## Measures of Goodness of Fit

When analyzing the effectiveness of a regression model, it is 
crucial to assess how well the model fits the data. Below we 
revisit the coefficient of determination $R^2$ and introduce 
additional measures.

### Coefficient of Determination {.unnumbered}

The **coefficient of determination** or $R^2$ is the percent of 
the variation in $y$ that is explained by changes in $x$. The 
higher the $R^2$ the better the explanatory power of the model.

::: {.callout-formula}
**THE COEFFICIENT OF DETERMINATION**
$$R^2=\frac{SSR}{SST}$$

where:

- $SST = \sum(y_i-\bar{y})^2$ — total variation in $y$
- $SSR = \sum(\hat{y_i}-\bar{y})^2$ — variation explained by the model
- $SSE = \sum(y_i-\hat{y_i})^2$ — variation unexplained by the model
- $SST = SSR + SSE$

$R^2$ is always between $[0,1]$.
:::

::: {.callout-example}
**Example:** Consider data on the Weight ($y$) and Exercise ($x$) 
of a particular person:

| Weight (y) | Exercise (x) |
|:----------:|:------------:|
|    165     |      45      |
|    170     |      10      |
|    168     |      25      |
|    164     |      30      |
|    165     |      40      |

If we used only the mean $\bar{y}=166.4$ as our prediction, the 
total squared error would be:

$$SST=(-1.4)^2+(3.6)^2+(1.6)^2+(-2.4)^2+(-1.4)^2=25.2$$

Using the regression line reduces these errors. The mistakes made 
by the regression line are quantified by:

$$SSE=0.81+0.29+0.69+5.76+0.02=7.57$$

The improvement gained by using regression over the mean is:

$$SSR=5.29+9.4+0.59+0+2.35=17.63$$

Hence the coefficient of determination is:

$$R^2=\frac{SSR}{SST}=\frac{17.63}{25.2}\approx0.70$$

The regression line explains about $70\%$ of the variation in 
Weight using Exercise.
:::

## Multiple Regression

If we were to predict weight we could use several other variables 
to get a better prediction. **Multiple regression** is a technique 
used to predict a variable using more than one independent variable.

::: {.callout-formula}
**THE MULTIPLE REGRESSION LINE**
$$\hat{y_i}=\hat{\beta_0}+\hat{\beta_1}x_1+\hat{\beta_2}x_2+...+\hat{\beta_k}x_k$$
where $k$ is the total number of independent variables included 
in the model.
:::

::: {.callout-example}
**Example:** Let's consider an additional variable in our Weight 
and Exercise example. The table below includes information on 
Calories ($z$):

| Weight (y) | Exercise (x) | Calories (z) |
|:----------:|:------------:|:------------:|
|    165     |      45      |     1200     |
|    170     |      10      |     1260     |
|    168     |      25      |     1220     |
|    164     |      30      |     1180     |
|    165     |      40      |     1190     |

The multiple regression line estimated by computer software is:

$$\hat{y_i}=83.96-0.025x+0.069z$$

Two conclusions follow from this result. First, as exercise 
increases weight tends to go down, while more calories increases 
weight. Second, reducing calorie consumption by $1$ unit decreases 
weight by $0.069$ pounds, whereas one additional minute of exercise 
only reduces weight by $0.025$ pounds — making calorie reduction 
more effective per unit.
:::

### Anova {.unnumbered}

The Anova table helps us understand how well each variable in our 
regression model explains the dependent variable. It decomposes 
the $SSR$ by variable and tracks the remaining errors.

::: {.callout-example}
**Example:** For the Weight, Exercise, and Calories regression, 
the Anova table is:

|  Source   | Sum Squares |
|:---------:|:-----------:|
|     x     |    17.63    |
|     z     |    6.55     |
| Residuals |    1.01     |

Exercise explains $17.63$ of the total variation. Adding Calories 
reduces the remaining unexplained variation by another $6.55$, 
bringing the total $SSR$ to $24.18$ and the $SSE$ down to $1.01$. 
The $R^2$ increases from $0.70$ to:

$$R^2=\frac{17.63+6.55}{25.2}=0.96$$
:::

### Adjusted $R^2$ {.unnumbered}

The **adjusted** $R^2$ penalizes a model for including additional 
explanatory variables, preventing artificial inflation of $R^2$.

::: {.callout-formula}
**THE ADJUSTED $R^2$**
$$\bar{R}^2=1-(1-R^2)\frac{n-1}{n-k-1}$$
where $k$ is the number of explanatory variables and $n$ is the 
sample size.
:::

::: {.callout-example}
**Example:** For the Weight example with both Exercise and Calories 
($k=2$, $n=5$, $R^2=0.96$):

$$\bar{R}^2=1-(1-0.96)\frac{5-1}{5-2-1}=0.92$$
:::

### Residual Standard Error {.unnumbered}

The **Residual Standard Error** estimates the average dispersion 
of the data points around the regression line.

::: {.callout-formula}
**THE RESIDUAL STANDARD ERROR**
$$s_e=\sqrt{\frac{SSE}{n-k-1}}$$
where $k$ is the number of independent variables and $n$ is the 
sample size.
:::

::: {.callout-example}
**Example:** For the Weight and Exercise model ($k=1$):

$$s_e=\sqrt{\frac{7.57}{5-1-1}}=1.59$$

Adding Calories ($k=2$) reduces this to:

$$s_e=\sqrt{\frac{1.01}{5-2-1}}=0.71$$

The smaller residual standard error confirms that adding Calories 
improves the model's fit.
:::

## Regression in Excel

Excel provides a straightforward way to estimate regression models 
using the **Data Analysis ToolPak**. If it is not already enabled, 
go to `File → Options → Add-ins → Analysis ToolPak → Go` and check 
the box.

[{{< fa file-excel >}} Download Data](weight_exercise.xlsx){.btn .btn-success}

**Simple Regression (Weight and Exercise):**

1. Go to `Data → Data Analysis → Regression`
2. Set the **Input Y Range** to the Weight column
3. Set the **Input X Range** to the Exercise column
4. Check **Labels** if your data includes headers
5. Select an output location and click `OK`

Excel will produce a full regression output including the 
coefficients, $R^2$, Adjusted $R^2$, Standard Error, and Anova 
table — all in one step.

**Reading the Excel Output:**

- **Intercept** row → $\hat{\alpha}$
- **X Variable 1** row → $\hat{\beta}$
- **R Square** → $R^2$
- **Adjusted R Square** → $\bar{R}^2$
- **Standard Error** → $s_e$

**Multiple Regression (Weight, Exercise, and Calories):**

Follow the same steps but set the **Input X Range** to include 
both the Exercise and Calories columns. Excel handles multiple 
predictors automatically.

**Making Predictions:** Once you have the coefficients from the 
output, use them directly in a formula. For the Advertisement 
and Price example, if the intercept is in cell B17 and slope 
in B18:

```excel
=B17 + B18 * 6
```

This returns the predicted price for an advertisement budget of $6$.

## Excel Function Summary

Below is a list of the Excel functions and tools used in this 
section:

- **Data Analysis ToolPak → Regression** is the primary tool for 
  estimating regression models in Excel. It returns coefficients, 
  $R^2$, adjusted $R^2$, standard error, and the Anova table in 
  a single output.

- `=SLOPE(y_range, x_range)` calculates the slope $\hat{\beta}$ 
  of the simple regression line directly.

- `=INTERCEPT(y_range, x_range)` calculates the intercept 
  $\hat{\alpha}$ of the simple regression line directly.

- `=FORECAST.LINEAR(x, y_range, x_range)` returns the predicted 
  value of $y$ for a given value of $x$ using simple linear 
  regression.

- `=RSQ(y_range, x_range)` returns the $R^2$ for a simple 
  regression.

- `=STEYX(y_range, x_range)` returns the residual standard error 
  $s_e$ for a simple regression.

## Exercises

The following exercises will help you get practice on Regression 
Line estimation and interpretation. In particular, the exercises 
work on:

- Estimating the slope and intercept.
- Calculating measures of goodness of fit.
- Prediction using the regression line.

Answers are provided below. Try not to peak until you have 
formulated your own answer and double checked your work for 
any mistakes.

### Exercise 1 {.unnumbered}

For the following exercises, make your calculations by hand and 
verify results using R functions when possible.

1.  Consider the data below. Calculate the deviations from the 
mean for each variable and use the results to estimate the 
regression line. Use R to verify your result. On average by how 
much does *y* increase per unit increase of *x*?

| **x** | 20  | 21  | 15  | 18  | 25  |
|:-----:|:---:|:---:|:---:|:---:|:---:|
| **y** | 17  | 19  | 12  | 13  | 22  |

<details>
<summary>Answer</summary>

*The regression lines is* $\hat{y}=-4.93+1.09x$. For each unit 
increase in x, y increases on average $1.09$.

*Start by generating the deviations from the mean for each 
variable. For x the deviations are:*

```{r}
x<-c(20,21,15,18,25)
(devx<-x-mean(x))
```

*Next, find the deviations for y:*

```{r}
y<-c(17,19,12,13,22)
(devy<-y-mean(y))
```

*For the slope we need to find the deviation squared of the x's. 
This can easily be done in R:*

```{r}
(devx2<-devx^2)
```

*The slope is calculated by* 
$\frac{\sum_{i=i}^{n}(x_{i}-\bar{x})(y_{i}-\bar{y})}{\sum_{i=i}^{n}(x_{i}-\bar{x})^2}$. 
In R we can just find the ratio between the summations of 
(devx)(devy) and devx2.

```{r}
(slope<-sum(devx*devy)/sum(devx2))
```

*The intercept is given by* $\bar{y}-\beta(\bar{x})$. In R we 
find that the intercept is equal to:

```{r}
(intercept<-mean(y)-slope*mean(x))
```

*Our results can be easily verified by using the `lm()` and 
`coef()` functions in R.*

```{r}
fitEx1<-lm(y~x)
coef(fitEx1)
```

</details>

2.  Calculate *SST*, *SSR*, and *SSE*. Confirm your results in R. 
What is the $R^2$? What is the Standard Error estimate? Is the 
regression line a good fit for the data?

<details>
<summary>Answer</summary>

*SST is* $69.2$, SSR is $64.82$ and SSE is $4.38$ (note that 
$SSR+SSE=SST$). The $R^2$ is just $\frac{SSR}{SST}=0.94$ and 
the Standard Error estimate is $1.21$. They both indicate a great 
fit of the regression line to the data.

*Let's start by calculating the SST. This is just* 
$\sum{(y_{i}-\bar{y})^2}$.

```{r}
(SST<-sum((y-mean(y))^2))
```

*Next, we can calculate SSR. This is calculated by the following 
formula* $\sum{(\hat{y_{i}}-\bar{y})^2}$. To obtain the predicted 
values in R, we can use the output of the `lm()` function. Recall 
our fitEx1 object created in Exercise 1. It has fitted.values 
included:

```{r}
(SSR<-sum((fitEx1$fitted.values-mean(y))^2))
```

*The ratio of SSR to SST is the* $R^2$:

```{r}
(R2<-SSR/SST)
```

*Finally, let's calculate SSE* $\sum{(y_{i}-\hat{y_{i}})^2}$:

```{r}
(SSE<-sum((y-fitEx1$fitted.values)^2))
```

*With the SSE we can calculate the Standard Error estimate:*

```{r}
sqrt(SSE/3)
```

*We can confirm these results using the `summary()` function.*

```{r}
summary(fitEx1)
```

</details>

3.  Assume that *x* is observed to be *32*, what is your 
prediction of *y*? How confident are you in this prediction?

<details>
<summary>Answer</summary>

*If* $x=32$ then $\hat{y}=29.87$. The regression is a good fit, 
so we can feel good about our prediction. However, we would be 
concerned about the sample size of the data.

*In R we can obtain a prediction by using the `predict()` 
function. This function requires a data frame as an input for 
new data.*

```{r}
predict(fitEx1, newdata = data.frame(x=c(32)))
```

</details>

### Exercise 2 {.unnumbered}

You will need the **Education** data set to answer this question. 
You can find the data set at 
https://jagelves.github.io/Data/Education.csv . The data shows 
the years of education (*Education*), and annual salary in 
thousands (*Salary*) for a sample of $100$ people.

1.  Estimate the regression line using R. By how much does an 
extra year of education increase the annual salary on average? 
What is the salary of someone without any education?

<details>
<summary>Answer</summary>

*An extra year of education increases the annual salary about* 
$5,300$ dollars (slope). A person that has no education would 
be expected to earn $17,258$ dollars (intercept).

*Start by loading the data in R:*

```{r message=FALSE}
library(tidyverse)
Education<-read_csv("https://jagelves.github.io/Data/Education.csv")
```

*Next, let's use the `lm()` function to estimate the regression 
line and obtain the coefficients:*

```{r}
fitEducation<-lm(Salary~Education, data = Education)
coefficients(fitEducation)
```

</details>

2.  Confirm that the regression line is a good fit for the data. 
What is the estimated salary of a person with $16$ years of 
education?

<details>
<summary>Answer</summary>

*The* $R^2$ is $0.668$ and the standard error is $21$. The line 
is a moderately good fit. If someone has $16$ years of education, 
the regression line would predict a salary of $102,000$ dollars.

*Let's get the* $R^2$ and the Standard Error estimate by using 
the `summary()` function and fitEducation object.

```{r}
summary(fitEducation)
```

*Lastly, let's use the regression line to predict the salary for 
someone who has* $16$ years of education.

```{r}
predict(fitEducation, newdata = data.frame(Education=c(16)))
```

</details>

### Exercise 3 {.unnumbered}

You will need the **FoodSpend** data set to answer this question. 
You can find this data set at 
https://jagelves.github.io/Data/FoodSpend.csv .

1.  Omit any NA's that the data has. Create a dummy variable that 
is equal to $1$ if an individual owns a home and $0$ if the 
individual doesn't. Find the mean of your dummy variable. What 
proportion of the sample owns a home?

<details>
<summary>Answer</summary>

*Approximately,* $36$% of the sample owns a home.

*Start by loading the data into R and removing all NA's:*

```{r}
Spend<-read_csv("https://jagelves.github.io/Data/FoodSpend.csv")
Spend<-na.omit(Spend)
```

*To create a dummy variable for OwnHome we can use the `ifelse()` 
function:*

```{r}
Spend$dummyOH<-ifelse(Spend$OwnHome=="Yes",1,0)
```

*The average of the dummy variable is given by:*

```{r}
mean(Spend$dummyOH)
```

</details>

2.  Run a regression with *Food* being the dependent variable and 
your dummy variable as the independent variable. What is the 
interpretation of the intercept and slope?

<details>
<summary>Answer</summary>

*The intercept is the average food expenditure of individuals 
without homes (*$6417$). The slope is the difference in food 
expenditures between individuals that do have homes minus those 
who don't. Individuals that do have a home spend about $-2516$ 
less on food than those who don't have homes.

*To run the regression use the `lm()` function:*

```{r}
lm(Food~dummyOH,data=Spend)
```

</details>

3.  Now run a regression with *Food* being the independent 
variable and your dummy variable as the dependent variable. What 
is the interpretation of the intercept and slope? Hint: you might 
want to plot the scatter diagram and the regression line.

<details>
<summary>Answer</summary>

*The scatter plot shows that most of the points for home owners 
are below* $6000$. For non-home owners they are mainly above 
$6000$. The line can be used to predict the likelihood of owning 
a home given someone's food expenditure. The slope tells us how 
that likelihood changes as the food expenditure increases by 1 — 
in general, the likelihood of owning a home decreases as food 
expenditure increases.

*Run the `lm()` function once again:*

```{r}
fitFood<-lm(dummyOH~Food,data=Spend)
coefficients(fitFood)
```

*For the scatter plot use the following code:*

```{r}
library(ggthemes)
Spend %>% ggplot() + 
  geom_point(aes(y=dummyOH,x=Food), 
             col="black", pch=21, bg="grey") +
  geom_smooth(aes(y=dummyOH,x=Food), method="lm",
              formula=y~x, se=F) + 
  theme_clean()
```

</details>

### Exercise 4 {.unnumbered}

You will need the **Population** data set to answer this question. 
You can find this data set at 
https://jagelves.github.io/Data/Population.csv .

1.  Run a regression of *Population* on *Year*. How well does 
the regression line fit the data?

<details>
<summary>Answer</summary>

*If we follow the* $R^2=0.81$ the model fits the data very well.

*Let's load the data from the web:*

```{r}
Population<-read_csv("https://jagelves.github.io/Data/Population.csv")
```

*Now let's filter the data so that we can focus on the population 
for Japan.*

```{r message=FALSE}
Japan<-filter(Population,Country.Name=="Japan")
```

*Next, we can run the regression of Population against the Year. 
Let's also run the `summary()` function to obtain the fit and 
the coefficients.*

```{r}
fit<-lm(Population~Year,data=Japan)
summary(fit)
```

</details>

2.  Create a prediction for Japan's population in 2030. What is 
your prediction?

<details>
<summary>Answer</summary>

*The prediction for* $2030$ is about $140$ million people.

*Let's use the `predict()` function:*

```{r}
predict(fit,newdata=data.frame(Year=c(2030)))
```

</details>

3.  Create a scatter diagram and include the regression line. 
How confident are you of your prediction after looking at the 
diagram?

<details>
<summary>Answer</summary>

*After looking at the scatter plot, it seems unlikely that the 
population in Japan will hit* $140$ million. Population has been 
decreasing in Japan!

*Use `ggplot` to create the figure.*

```{r}
Japan %>% ggplot() +
  geom_point(aes(y=Population,x=Year), 
             col="black", pch=21, bg="grey") +
  geom_smooth(aes(y=Population,x=Year), 
              formula=y~x, method="lm", se=F) +
  theme_clean()
```

</details>